The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2022-11-17 manuscript.qmd
Update lab notes

pagetitle: Cross-linguistic similarity and  word acquisition
abstract: |
    Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those form the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 10 to 36 months.
thanks: |
    This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances.
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
number-sections: true
fig-dpi: 1000
csl: apa.csl
bibliography: manuscript/references.bib
google-scholar: true
format:
  pdf:
    pdf-engine: xelatex
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    keep-tex: true
    linestretch: 1.25
---

@@

# load and process objects
tar_load(participants)
tar_load(items)
tar_load(multilex_data)
tar_load(model_log_liks)
tar_load(model_loos)
tar_load(model_fit_1)
tar_load(rope_coefs)
tar_load(rope_interval)
tar_load(df)
tar_load(posterior_description)

@@

# Introduction {#introduction}

The foundations of word learning are in place early in age: before the end of their first year of life infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999some; @bergelson2012months; @bergelson2015early] and parental reports [e.g., @fenson2007macarthur]. During the last half of their second year, they acquire new words at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002children; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation across children, reflected, for instance, on the variation on the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of those words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure, bates1994developmental].

@@

There is evidence that bilinguals know, on average, less words in each of their languages than monolinguals, but also that both groups know a similar amount of words when the two languages are aggregated. For example, @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, pointing to the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar--or even larger--number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms--one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in varying degrees. Particularly, when both languages are typologically close, like Spanish and Catalan (both Roman languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two linguistically distant languages, like Spanish and English (one Roman, the other Germanic). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning a pair of languages that share a high proportion of cognates should benefit more often from this facilitation effect than those learning a pair of languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@@

Floccia et al. pointed to *parallel activation* as the main mechanism underpinning their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest examples of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals--unfamiliar with the Catalan translations of the Spanish words they uttered--did not show this effect. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Several subsequent studies have also provided similar evidence in comprehension in children [e.g., @von2012language; @poulin2013lexical]. Parallel activation is therefore a plausible mechanism to account for Floccia et al.'s results: cognates increase the amount of activation in both languages, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. A missing piece in this account is how increase parallel activation might facilitate the acquisition new words.

In this study, we hypothesise that cognates are acquired earlier in age than non-cognates, which might explain why children learning languages that share many cognates show larger vocabulary sizes. Evidence supporting this claim is scarce. @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (much larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

@@

As @mitchell2022cognates note in their discussion, two competing hypotheses remain untested. The facilitative effect of cognateness can be explained by, at least, two alternative, arguably incompatible, mechanisms. On the one hand, cognateness might operate as a facilitator of the acquisition of both forms of a translation pair, with every exposure to either form activating the other, and therefore strengthening the association between both and their referent. If this is the case, then cognateness should increase the probability of acquisition of the forms in both languages equally.

Another possibility is that cognateness plays a role in the acquisition of a translation pair only when at least one of the word-forms has been acquired. This account departs from the premise that, if the child is not previously familiar with one of the word forms, the acquisition of its translation in the other language cannot be facilitated by cognateness. This implies that, if cognateness facilitates the acquisition of a word form, it can only do so after the child has acquired one of the word forms of the translation pair. Given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014much; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness.

In line with this second hypothesis, the vocabulary size associated with linguistic similarity that Floccia et al. reported was larger in the additional language vocabulary than in English vocabulary. Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have, on average, learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  If this is the case, then the acquisition of English words by English-dominant participants would rarely benefit from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

In this study, we investigated the role of cognateness on the acquisition of translation equivalent, and whether such effect is conditional to the amount of time the child is exposed to the corresponding language of each member of the translation pair. Using an online vocabulary checklist--designed specifically for this study--we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We then adopted an Item Response Theory (*IRT*) approach to model the probability of participants being reported by their parents to understand or understand *and* say each word in the checklist, conditional to its cognate status in Catalan and Spanish and participants' degree of exposure to the corresponding language, while adjusting for other indices of word difficulty--lexical frequency and length (phonemes)-- and participant skill--age [see @kachergis2022toward for a similar approach].

# Methods {#methods}

Data and materials are available at OSF [https://osf.io/hy984/](https://osf.io/hy984/), and code is available at GitHub [https://github.com/gongcastro/trajectories](https://github.com/gongcastro/trajectories).

## Participants {#participants}

@@

## Questionnaire {#questionnaire}

@@

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. In case two translation equivalents were possible for a given word, both were included as separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single item (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We included items from a diverse sample of `r n_categories` semantic/functional categories.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (`r format(n_te, big.mark = ",")` translation equivalents; `r format(n_item_language["catalan"], big.mark = ",")` items in Catalan, `r format(n_item_language["spanish"], big.mark = ",")` items in Spanish), while others filled a shorter version (~400 translation equivalents, ~400 items in Catalan, ~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were designed so that each contained a representative sub-sample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. For the analyses in the present study, we considered responses to words corresponding to nouns, verbs, and adjectives [@fourtassi2020growth]. We excluded multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono / mico*).

We extracted four variables of interest for each word-form: (1) Word lexical frequency ($Frequency$): lexical frequency of the word in its corresponding language, expressed as Zipf scores [@van2014subtlex; @zipf1949human]. This variable ranges from 0 to 7, and follows and approximates a normal distribution, with most values in a corpus ranging from 3 to 5 points. Lexical frequencies were extracted from an adult subtitle-based database: SUBTLEX-CAT [@boada2020subtlex] for Catalan words and SUBTLEX-ESP [@cuetos2012subtlex] for Spanish words. Using child-based estimates of lexical frequency like CHILDES was not possible due to the low number of Catalan participants and tokens in the available corpora. Responses to words with missing lexical frequency score were excluded from analyses (`r percent(n_missing_freq, accuracy = 0.01)` of the total number of items). (2) Word length ($Phonemes$): number of phonemes in the phonological transcription of the word-form in International Phonological association format. (3) Phonological similarity ($Levenshtein$): normalised Levenshtein similarity between the word-form and its translation in the other language [@levenshtein1966binary]. This score is calculated by first calculating the Levenshtein distance between the two transcriptions (number of insertions, deletions or replacements needed for the shortest transcription to become identical to the longer transcription), then dividing the resulting value by the length of the longest transcription, and finally subtracting this value from 1. This results in a proportion that indicates how much the two phonological transcriptions of the translation equivalent are similar to each other, ranging from 0% (no similarity at all) to 100% (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches]. (4) Participant age: number of months elapsed between participants' birth date and questionnaire completion. Finally, (5) participant degree of exposure to the words' language ($DoE$): percentage of exposure to the language the word belonged to. For example, for a participant with 90% exposure to Catalan, and 10% to Spanish, the DoE of the Catalan word *taula* would be 90%, and the DoE for the Spanish word *mesa* would be 10%. @tbl-item-summary shows a summary of the distribution of the word-level predictors across languages in and questionnaire versions.

```{r}
#| label: tbl-item-summary
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Lexical properties of the word-forms and translationpairs included in the vocabulary checklist and our analyses. We report the number of items, mean (*M*), standard deviation (*SD*), and range of the lexical frequency (epxressed in Zipf scores), length (number of phonemes) of the word forms, and the phonological similarity (normalised Levenshtein similarity) with their translation equivalent. Statistics are reported separately for Catalan and Spanish, and for the different vocabulary checklists that participants were allocated to (A, B, C, D)."
items <- left_join(items, select(multilex_data$pool, te, item, language, class, category))

items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq, n_phon, lv), 
        lst(n = ~sum(!is.na(.)), mean, sd, min, max)
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_n) %>% 
    gt(groupname_col = "language") %>%
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("# Phonemes", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Phon. similarity (Levenshtein)", columns = starts_with("lv_")) %>% 
    cols_merge_range(col_begin = "freq_min", col_end = "freq_max") %>% 
    cols_merge_range(col_begin = "n_phon_min", col_end = "n_phon_max") %>% 
    cols_merge_range(col_begin = "lv_min", col_end = "lv_max") %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_mean = md("Mean"),
        freq_sd = md("*SD*"),
        freq_min = md("Range"),
        n_phon_mean = md("Mean"),
        n_phon_sd = md("*SD*"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean"),
        lv_sd = md("*SD*"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    )
```

## Data analysis {#analysis}

@@

n_te <- length(unique(df$te))
```

@@

post_draws_list <- split(posterior_description, posterior_description$parameter)

# summarise posterior draws
posterior_description %>% 
    mutate(
        parameter = factor(
            parameter,
            levels = names(str_repl),
            labels = str_repl
        ) %>% 
            as.character()
    ) %>%
    select(parameter, median, ci_high, ci_low, rope_percentage) %>%  
    gt() %>% 
    fmt_number(2:4) %>%
    fmt_percent(5) %>%
    cols_merge(c(ci_low, ci_high), pattern = "[{1}, {2}]") %>%
    cols_label(
        parameter = "Parameter",
        median = "Median",
        ci_low = md("95% *HDI*"),
        rope_percentage = md("*ROPE* prob.")
    ) %>%
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = 1:5)
    )
```

>>>>

pagetitle: Cross-linguistic similarity and  word acquisition
thanks: |
    This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances.
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
number-sections: true
fig-dpi: 1000
csl: apa.csl
bibliography: manuscript/references.bib
google-scholar: true
format:
  pdf:
    pdf-engine: xelatex
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    keep-tex: true
    linestretch: 1.25
  docx:
    reference-doc: manuscript/manuscript_template.docx
    linestretch: 1.25
---

@@

# load and process objects
tar_load(participants)
tar_load(items)
tar_load(multilex_data)
tar_load(model_log_liks)
tar_load(model_loos)
tar_load(model_fit_0)
tar_load(model_fit_1)
tar_load(model_fit_2)
tar_load(model_fit_4)
tar_load(rope_coefs)
tar_load(rope_interval)
tar_load(df)
tar_load(posterior_description)

@@

# Introduction {#sec-introduction}

## Word acquisition during toddlerhood: the effect of language experience

The foundations of word learning are in place early in age. Before the end of their first year of life infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999some; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthur]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002children; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation across children. This is reflected, for instance, on the variability of the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of acquired words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure, @bates1994developmental].

@@

Previous findings point to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, but also to both groups knowing a similar amount of words when the two languages are aggregated. For example, @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, highlighting the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar--or even larger--number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

## Language distance as a mediator of bilingual vocabulary growth

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms--one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in varying degrees. Particularly, when both languages are typologically close, like Spanish and Catalan (both Roman languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically distant languages, like Spanish and English (one Romance, the other Germanic). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning a pair of languages that share a high proportion of cognates should benefit more often from this facilitation effect than those learning a pair of languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@@

## Lexical non-selectiviy as a candidate acount: the role of parallel activation

Floccia et al. pointed to *parallel activation* as the main phenomenon underpinning their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest examples of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals--unfamiliar with the Catalan translations of the Spanish words they uttered--did not show this effect. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Several subsequent studies have also provided similar evidence in comprehension in children [e.g., @von2012language; @poulin2013lexical]. Parallel activation is therefore a plausible explanation for Floccia et al.'s results: cognateness increases the amount of cross-language activation, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. This account is is line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words, already acquired or not [e.g., @hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. Importantly, bilinguals seem to be more likely to acquire words for which their translation equivalent has been already learnt in the other language, suggesting that semantic similarity also facilitates word acquisition across languages [@bilson2015semantic].

If phonological similarity also plays a cross-language facilitation role during word acquisition, cognate translation equivalents--which share both semantic and phonological similarity--should be acquired earlier that non-cognate translation equivalents--which share semantic similarity, but not phonological similarity. Evidence supporting and earlier age of acquisition for cognates is scarce. @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (much larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

@@

## Two competing mechanisms for a cognate facilitation during word acquisition

The facilitative effect of cognateness on word acquisition can be explained by at least two arguably incompatible mechanisms. On the one hand, and in line with @mitchell2022cognates, cognateness might operate at early stages of word acquisition, before stable representations of any of the word-forms of the translation equivalent have been formed. In this scenario, recognising any of the two word-forms as familiar might lead to the activation of its (cognate) translation in the other language via their phonological similarity. Given that both word-forms are likely to be heard in the presence of the same referent, the association between each form and such reference could be facilitated by the exposure to either of them.

Other words might be activated as well for the same reasons (similar sounding words in either language but with different meaning), but only the acquisition of translation equivalents might be f

very exposure to a word-form activates its (phonologically similar) translation, strengthening the association between both word-forms and their referent, even when only the form in one of the languages is heard most of the times.

For instance, hearing *flower* in English might activate *floeur* in French, and *vice versa*. Repeated exposure to the English or the French words, and therefore to similar -sounding words in the presence of  eventually leads to an earlier acquisition for both of them, as compared to non-cognate translation equivalents like *cow* and *vache*, which do not activate each other via phonological similarity. If this is the case, then cognateness should increase the probability of acquisition of the word-forms in both languages equally.

Another possibility is that cognateness plays a role in the acquisition of a translation pair only when at least one of the word-forms has been acquired. This account departs from the premise that phonological similarity operates at the lexical level to facilitate word learning across languages: if the child has still not formed a lexical representation with sufficient phonological detail for one of the word forms, the acquisition of its translation in the other language cannot be facilitated by cognateness. This implies that, if cognateness facilitates the acquisition of a word form, it can only do so after the child has acquired one of the word forms of the translation pair. Given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014much; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness.

In line with this second hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure. Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have, on average, learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  If this is the case, then the acquisition of English words by English-dominant participants would rarely benefit from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

## The present study

In this study, we investigated the role of cognateness on the acquisition of translation equivalents, and if such effect is conditional to the amount of time the child is exposed to the corresponding language of each member of the translation pair. Using an online vocabulary checklist--designed specifically for this study--we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We then adopted an Bayesian item response theory (*IRT*) approach to model the probability of participants being reported by their parents to understand or understand *and* say each word in the checklist, conditional to its cognate status in Catalan and Spanish and participants' degree of exposure to the corresponding language, while adjusting for other indices of word difficulty--lexical frequency [@goodman2008does] and length (phonemes)-- and participant skill--age [see @kachergis2022toward for a similar approach].

# Methods {#sec-methods}

## Participants {#sec-participants}

@@

## Questionnaire {#sec-questionnaire}

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire, MultiLex [@multilex]. This questionnaire was inspired in the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthur] and its adaptations to other languages, and was implemented on-line using the formR platform [@arslan2020formr]. This questionnaire is structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. In case two translation equivalents were possible for a given word, both were included as separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single item (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We included items from a diverse sample of `r n_categories` semantic/functional categories.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (`r format(n_te, big.mark = ",")` translation equivalents; `r format(n_item_language["catalan"], big.mark = ",")` items in Catalan, `r format(n_item_language["spanish"], big.mark = ",")` items in Spanish), while others filled a shorter version (~400 translation equivalents, ~400 items in Catalan, ~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were designed so that each contained a representative sub-sample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. For the analyses in the present study, we considered responses to words corresponding to nouns, verbs, and adjectives [@fourtassi2020growth]. We excluded multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono/mico*).

## Variables of interest

We extracted four variables of interest for each word-form. (1) Word lexical frequency ($Frequency$): lexical frequency of the word in its corresponding language, expressed as Zipf scores [@van2014subtlex; @zipf1949human]. This variable ranges from 0 to 7, and follows and approximates a normal distribution, with most values in a corpus ranging from 3 to 5 points. Lexical frequencies were extracted from an adult subtitle-based database: SUBTLEX-CAT [@boada2020subtlex] for Catalan words and SUBTLEX-ESP [@cuetos2012subtlex] for Spanish words. Using child-based estimates of lexical frequency like CHILDES was not possible due to the low number of Catalan participants and tokens in the available corpora. Responses to words with missing lexical frequency score were excluded from analyses (`r percent(n_missing_freq, accuracy = 0.01)` of the total number of items). (2) Word length ($Phonemes$): number of phonemes in the phonological transcription of the word-form in International Phonological association format. (3) Phonological similarity ($Levenshtein$): normalised Levenshtein similarity between the word-form and its translation in the other language [@levenshtein1966binary]. This score is calculated by first calculating the Levenshtein distance between the two transcriptions (number of insertions, deletions or replacements needed for the shortest transcription to become identical to the longer transcription), then dividing the resulting value by the length of the longest transcription, and finally subtracting this value from 1. This results in a proportion that indicates how much the two phonological transcriptions of the translation equivalent are similar to each other, ranging from 0% (no similarity at all) to 100% (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches]. (4) Participant age: number of months elapsed between participants' birth date and questionnaire completion. Finally, (5) participant degree of exposure to the words' language ($DoE$): percentage of exposure to the language the word belonged to. For example, for a participant with 90% exposure to Catalan, and 10% to Spanish, the DoE of the Catalan word *taula* would be 90%, and the DoE for the Spanish word *mesa* would be 10%. @tbl-item-summary shows a summary of the distribution of the word-level predictors across languages in and questionnaire versions.

```{r}
#| label: tbl-item-summary
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Lexical properties of the word-forms and translation pairs included in the vocabulary checklist and our analyses. We report the number of items, mean, standard deviation (*SD*), and range of the lexical frequency (epxressed in Zipf scores), length (number of phonemes) of the word forms, and the phonological similarity (normalised Levenshtein similarity) with their translation equivalent. Statistics are reported separately for Catalan and Spanish, and for the different vocabulary checklists that participants were allocated to (A, B, C, D)."
items <- left_join(items, select(multilex_data$pool, te, item, language, class, category))

items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq, n_phon, lv), 
        lst(n = ~sum(!is.na(.)), mean, sd, min, max)
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_n) %>% 
    gt(groupname_col = "language") %>%
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("Length (Phonemes)", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Phon. similarity (Levenshtein)", columns = starts_with("lv_")) %>% 
    cols_merge_range(col_begin = "freq_min", col_end = "freq_max") %>% 
    cols_merge_range(col_begin = "n_phon_min", col_end = "n_phon_max") %>% 
    cols_merge_range(col_begin = "lv_min", col_end = "lv_max") %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_mean = md("Mean"),
        freq_sd = md("*SD*"),
        freq_min = md("Range"),
        n_phon_mean = md("Mean"),
        n_phon_sd = md("*SD*"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean"),
        lv_sd = md("*SD*"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    )
```

## Data analysis {#sec-analysis}

@@

n_te <- length(unique(df$te))

rope_interval <- c(lower = -0.1, upper = 0.1)
```

@@

posterior_description <- bayestestR::describe_posterior(
    model_fit_4,
    test = "rope",
    rope_range = rope_interval, 
    ci_method = "HDI"
) %>% 
    as_tibble() %>% 
    clean_names()

post_draws_list <- split(posterior_description, posterior_description$parameter)
```
